---
layout: post
title: "Send an email to all your representatives. With R."
categories: [hack]
tags: [R, XML]
date: 2015-07-15
description: Script to scramble the website of the french senat and send 300 emails. 
photo_url: senat.jpg
---

I disagreed with the recent law the French government pass.

I wanted to explain my point of view to the French senators and did it manually at first. After sending two mails, it hits me that it is possible to do a script to automate it and even a R script.

The two main packages used are `XML` and `mailR`, respectively to do the HTML scrambling and automatically sent mails.

The XML package is really simple to handle for this task and have the `getHTMLLinks()` function which extract all the link from a webpage.

<h1> Scrambling the senate website </h1>

In the French senate website, we could find a personal page for each senator and a few information on our representatives, including the group, the mail provided by the senate and sometime personals mails.

The first step is to get all the personal web pages of the senators.

```{r, eval = F}

# packages used: 
library(XML) # install.packages("XML")
library(plyr) # install.packages("plyr")
library(stringr) # install.packages("stringr")

# url of the main page of the senat website
xml.url.short <- "http://www.senat.fr" 
xml.url <- paste0(xml.url.short, "/senateurs/senatl.html")

### get all the links of the senator website:
senateur.raw.link <- getHTMLLinks(xml.url)

## cleaning the set of link:
  # we keep only the ones beginning by /senateur/:
senateur.link <- unlist(senateur.raw.link)[substr(unlist(senateur.raw.link), 1, 10) == "/senateur/"]
  # we add the full path:
senateur.full.link <- paste0(xml.url.short, senateur.link)

### now, getting every mail address from these link:
  # function to get: name, surname, mail address, group, profession, circonscription
      # based on the link
senateur.infos.primaire <- data.frame(link = senateur.full.link
                , mail = "", full.name.surname = "", group = "", stringsAsFactors = F)

# timer
ptm <- proc.time()

for (i in 1:length(senateur.full.link)) { # i <- 1
  
  # mail:
  links <- unlist(getHTMLLinks(senateur.full.link[i], externalOnly = F))
  mails <- links[substr(links, 1, 6) == "mailto"]
  mail <- substr(mails[!substr(mails, 1, 20) %in% c("mailto:notices-senat", "mailto:?subject=&bod")], 8, 200)
  senateur.infos.primaire$mail[i] <- ifelse(length(mail) != 0, mail[length(mail)], "")
  
  # name and surname:
senateur.infos.primaire$full.name.surname[i] <- substr(mails[substr(mails, 1, 20) %in% c("mailto:notices-senat")], 66, 200)
}

print(ptm0 <- ptm - proc.time())

## convert to correct encodin to deal with the special characters
senateur.infos.primaire$full.name.surname <- iconv(senateur.infos.primaire$full.name.surname, from = "UTF-8", to = "LATIN2")

# sexe definition:
senateur.infos.primaire$sexe <- ifelse(substr(senateur.infos.primaire$full.name.surname, 1, 5) == "de la", "F", " ")
senateur.infos.primaire$sexe <- ifelse(substr(senateur.infos.primaire$full.name.surname, 1, 4) == "du S", "M", senateur.infos.primaire$sexe)

```


```{r, eval = F, include = F}

# name shortcut:
senateur.infos.primaire$name.shortcut <- ifelse(senateur.infos.primaire$sexe == "F", substr(senateur.infos.primaire$full.name.surname, 17, 200), substr(senateur.infos.primaire$full.name.surname, 13, 200) )

## names which will have to be dealt manually:
issue.name0 <- laply(senateur.infos.primaire$name.shortcut, function(x) ifelse(length(unlist(str_split(x, " "))) != 2, x, ""))
issue.name <- issue.name0[issue.name0 != ""]

# name & surname
senateur.infos.primaire$name0 <- laply(str_split(senateur.infos.primaire$name.shortcut, " "), function(x) x[1])
senateur.infos.primaire$surname0 <- laply(str_split(senateur.infos.primaire$name.shortcut, " "), function(x) x[2])

## manual split 
manual.name <- data.frame(issue.name = issue.name, name1 = c("del Picchia", "Des Esgaulx", "di Folco"
        , "Estrosi", "de Legge", "Le scouarnec", "Masson", "Mohamed Soilihi", "de Montgolfier", "de Nicolay"
        , "Pozzo di Borgo", "de Raincourt", "Vogel")
      , surname1 = c("Robert", "Marie-Helene", "Catherine", "Dominique", "Dominique", "Michel"
        , "Jean Louis", "Thani", "Alberic", "Louis-Jean", "Yves", "Henri", "Jean Pierre"), stringsAsFactors = F)

### merge:
senateur.infos <- merge(senateur.infos.primaire, manual.name, by.x = "name.shortcut", by.y = "issue.name", all.x = T)

## final version of the name & surname
senateur.infos$name <- ifelse(!is.na(senateur.infos$name1), senateur.infos$name1, senateur.infos$name0)
senateur.infos$surname <- ifelse(!is.na(senateur.infos$surname1), senateur.infos$surname1, senateur.infos$surname0)

write.csv(senateur.infos, file = "DATA/Raw_data_senateurs_info.csv")
save(senateur.infos, file = "DATA/Raw_data_senateurs_info.Rdata")

```

In the end, we get a file with all the emails of the French senators and the associated details.

<h1> Sending the mails </h1>

The package mailR allow to send email within R. It is very useful when you have really long process and you want to be averted when the process is finished.

The function send.mail is very practical, as it allows to send HTML formatted emails, which allow a great deal of customization.

To facilitate the debugging, the mail is written in HTML then the `brew` package is used to replace the names, and personalize the message looking at the group. It allows as well to keep the log of the mail.

```{r, eval = F}

## rJava is needed first
library(mailR)  # install.packages("mailR", dep = T)  
library(brew)  # install.packages("brew")
# install.packages("rJava", repos = "http://cran.univ-paris1.fr/") #  options(java.home="C:\\Program Files\\Java\\jre1.8.0_45\\") # library(rJava)

for(i in 1:dim(senateur.infos.primaire)[1]) {  #i <- 1

# create the log file
brew(file = paste0(path, "/Mail Template/template mail.html"), output = paste0(path, "/Log/Mail_", i, ".html") )

# the body of the mail
body <- paste0(scan(file = paste0(path, "/Log/Mail_", i, ".html"), what = "raw", sep = "|"), collapse = "")

from <- "my@email.com"
to <- senateur.infos.primaire[i, "mail"]

tt <- send.mail(from = from, to = to
          , subject = "Loi renseignement"
          , body = body
          , html = TRUE
          , smtp = list(host.name = "smtp.gmail.com", port = 465, user.name = "my username"
                      , passwd = "my password", ssl = TRUE)
          , authenticate = TRUE
          , SEND = TRUE)
}
```

The loop for sending the mail could take a few minutes.

<h1> Parity Analysis </h1>

The file allow an analysis of the parity. 

```{r, include = F}
# save(senateur.infos.primaire, file = paste0(path, "Raw_data_senateurs_info_primaire.Rdata"))
library(ggplot2)
path <- "C:/blog/gen/"
load(file = paste0(path, "data/Raw_data_senateurs_info.Rdata"))
senateur.infos <- read.csv(paste0(path, "data/SenatorList.csv"))
senateur.infos$sexe <- senateur.infos$genre
senateur.infos$group <- senateur.infos$Groupe
senateur.infos$group <- gsub("app. ", "", senateur.infos$group)
senateur.infos$group <- gsub("ratt. ", "", senateur.infos$group)

##ggsave("senateur_plot.png", senateur.plot, bg = "transparent", dpi = 500, width = 31.75, height = 10.58, units = "cm")

```

```{r, fig.width = 12}
# plot
ggplot(senateur.infos, aes(x = group, fill= sexe)) + geom_bar() +
  scale_x_discrete(name = "Group") +
  scale_y_continuous(name = "Number of senator\n") +
  scale_fill_discrete(name = "Sex") +
  ggtitle("Repartition of sex among the senators by group\n") +
  theme(axis.text.x = element_text(colour = "blue", size = 10)
        , axis.text.y = element_text(colour = "blue", size = 13)
        , panel.background = element_rect(fill = "white")
        , panel.grid.major.y = element_line(colour = "darkgrey", size = 1)
        , panel.grid.minor.y = element_line(colour = "grey", size = 0.5)
        , panel.grid.major.x = element_blank()
        , plot.title = element_text(size = 20, colour = "darkblue")
        , legend.title = element_text(size = 15, colour = "darkcyan")
        , legend.text = element_text(size = 13, colour = "blue")
        , axis.title = element_text(size = 15, colour = "darkcyan", face = "bold"))

```

Only the CRC (communists) and the ecologists have a strict parity.

